---
title: "02 Vectors"
output: 
  html_document:
    theme: dark
    highlight: zenburn
---

```{css, echo=FALSE}
/* This code is used to turn the html output into "dark mode" */
.dark-output {
background-color: #5a5a5a;
color: white;
}

.error-output {
background-color: #cc9393;
color: black;
}

.warning-output {
background-color: #c0c537;
color: black;
}

.message-output {
background-color: #5a5a5a;
color: white;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, class.output = "dark-output", class.error = "error-output",
  class.message = "message-output", class.warning = "warning-output"
)
```

## Vectors 

This is what a "vector" looks like:

```{r}
c(1, 7, 12, 4, 2)
```

You can see that this vector contains five numbers.
Those numbers are separated with commas and wrapped in parentheses - but at the beginning, there is an ominous `c`.
If you run this code, however, it just shows the numbers next to each other, without the `c()` and without the commas.

Here is another example of a vector:

```{r}
c(2, 6.1, 9.234, 1.23)
```

And another one:

```{r}
c("hello", "cake", "biscuit")
```

And another one:

```{r}
c(TRUE, FALSE, FALSE, TRUE)
```

We see that a vector is a collection of elements, and it can consist of all the data types we previously got to know.
There is only one rule: You can't mix data types in a vector.

This does not work:

```{r}
c(10, "biscuit", TRUE)
```

It **seems** like it worked, but the quotation marks around every element indicate that everything has just been turned into a character!

Look at what happens when we mix integers and logicals [^logical_numbers]:

```{r}
c(10, TRUE, FALSE)
```

[^logical_numbers]: Logicals have a numeric representation: `TRUE` equals 1, and `FALSE` equals 0.

You might also have noticed that in the second vector we looked at (`c(2, 6.1, 9.234, 1.23)`), all numbers were displayed with the same number of decimal places when you ran the code.
So beware - a vector forces all elements to be of the same data type!

Now for the mysterious `c`.
It's easy to understand why we wrap our vector in parentheses to define a group of elements, but why does the `c()` have to go in front of it?
`c()` is actually a function that combines elements (more about functions later).
In `R`, you need this function to be able to combine elements into a vector.
This will not work[^run_error]:

```{r error=TRUE}
(2, 5, 10)
```

[^run_error]: Note that an Rmarkdown file will not run if a chunk contains an error - unless you explicitly tell it that it's fine and that it is supposed to expect an error here. You can do this by setting the chunk option `error = TRUE` (see at the top of the chunk). If you ever want to purposely include an error in the code, e.g. for educational purposes, this is what you have to use.

In the beginning, you will probably forget about the `c()` a lot, but you'll get used to it soon.

Of course, vectors can be saved into variables, just like we did with single elements before:

```{r}
my_vector <- c(1, 4, 9, 12) # store vector in a variable
my_vector # print vector
```

1.1) Create some vectors. Create one vector with characters, one with numbers, and one with logicals. Save them all into different variables.

```{r}
coffee_break <- c("coffee", "cake", "biscuits")
favourite_numbers <- c(12, 3, 42, 666)
some_decimals <- c(1.11, 2.22, 3.33)
best_animal <- "owl"
```

1.2) We saw that the `c()` function combines elements into a vector, but it can also combine vectors! Try to combine some of the vectors you previously created using the `c()` function.

```{r}
c(coffee_break, best_animal)
```

### Optional

This isn't the first time you've used a vector. In fact, everything we did in "basics" exercises was about vectors. E.g., `2` is a one-element vector, just like `"hello"`. For one-element vectors, we don't need `c()`, because there isn't anything that needs to be combined.
We can check whether something is a vector using the function `is.vector()`. It returns `TRUE` when something is a vector.
This is how we can confirm that single elements are also a (very short) vector.
We will later get to know data types that are not a vector.

```{r}
is.vector(c(1, 2, 3))
is.vector("test")
is.vector(FALSE)
```

1.3) Actually, there was a vector in the "basics" exercises that had more than one element (but it didn't use `c()`). Can you find it?

## Mathematical operations

The important thing about vectors is that you can do mathematical operations with them.

2.1) Can you figure out what happens here?

```{r}
c(1, 2, 5) + 1
```

What do you expect the following code will produce?

2.2) `c(1, 2, 5) - 1`

```{r}
c(1, 2, 5) - 1
```

2.3) `c(1, 2, 5) * 2`

```{r}
c(1, 2, 5) * 2
```

2.4) `c(1, 2, 5) / 2`

```{r}
c(1, 2, 5) / 2
```

Vectors are one of the most important concepts of `R`.
Sure, other programming languages have similar concepts, but in `R`, vectors are really powerful.
A lot of things are "vectorised", that means that functions work on entire vectors, or that clever combinations of vectors can be used to produce the desired results.

As usual, it is easier to look at it than to describe it:

```{r}
c(2, 4, 6) + c(1, 0, 2)
```

Now, you can infer what these lines of code will do:

2.5) `c(2, 4, 6) - c(1, 0, 1)`

```{r}
c(2, 4, 6) - c(1, 0, 1)
```

2.6) `c(2, 4, 6) * c(1, 0, 1)`

```{r}
c(2, 4, 6) * c(1, 0, 1)
```

2.7) `c(2, 4, 6) / c(1, 2, 1)`

```{r}
c(2, 4, 6) / c(1, 2, 1)
```

However, things get *really* wild when we use vectors that don't have the same length.

Can you see what happened here?

```{r}
c(2, 4, 6, 5, 0, 0) + c(1, 60)
```

Let's look at this concept some more.

2.8) Can you multiply **every other** element of `c(2, 4, 6, 4, 9, 12)` with 2? (I.e. only 4, 4 and 12 should be multiplied with 2 - the other elements should be left unchanged.)

```{r}
c(2, 4, 6, 4, 9, 12) * c(1, 2, 1, 2, 1, 2)
c(2, 4, 6, 4, 9, 12) * c(1, 2)
```

### Optional

2.9) Given `c(2, 4, 6, 4, 9, 12)`, subtract 1 from the first element, 2 from the second, 3 from the third, then again 1 from the fourth, 2 from the second and so on.

2.10) Run the following code cell. What happened here? Is that bad?

```{r}
c(2, 4, 6, 4, 9, 12) + c(1, 2, 3, 4)
```

**THIS IS A GIT PUSH CHECKPOINT!**
