---
title: "04 Data Frames"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Most likely, the most frequent data format you will encounter during data analysis is the **data frame**.
It is more or less a table, similar to the data structure you would use in Excel or SPSS.

To explore data frames, we will use data from the "Nerdy Personality Attributes Scale", a fun scale that was developed to quantify what "nerdiness" is.
You can take the test online at [Open Psychometrics](https://openpsychometrics.org/tests/NPAS/).
The data for the test (and many others) is also freely available [here](https://openpsychometrics.org/_rawdata/).

The nerd data are stored in a .csv file - .csv and text files are really easy to read into `R`.
If we want to read in data from excel files or SPSS files, we need to install additional packages.
The function `read.csv()` does exactly what the name suggests: It reads in a .csv file.
Just like with the `load()` function in the previous chapter, it takes the path of the file we want to read in as its first argument.
Once again, you can find the data in the folder `"data_sets"`, under the name `"nerd_data.csv"`.

SWe try to read in our data and assign it to a variable called `nerd`.
This code runs fine, but ...

```{r}
nerd <- read.csv("exercises_solutions/data_sets/nerd_data_short.csv")

# Should the paths above return an error for you, try going to
# Tools -> Global Options -> R Markdown
# and set "evaluate chunks in directory:" to "Project"
```

... when we inspect the data using the `head()` function, it looks very weird:

```{r}
head(nerd)
```

If we look into our environment tab in the top right corner, it says that `nerd` has 1,000 observations (i.e. rows) of 1 variable - i.e. 1 column.
This is not the nice data table we expected.
That's because `R` didn't know how to separate the data into columns.
It put all of the information it found into one single column.
By default, `read.csv()` thinks the data are separated by a comma, (see `?read.csv`), but in our case, the columns are separated by a tab (indicated by a `\t`).
So, we need to tell `read.csv()` what the correct separator is via the argument `sep`.[^sep]

[^sep]: `read.csv()` is just a special case of the function `read.table()`. In principle, `read.table()` can be used to read in any form of data (table), as long as you specify a couple of things, e.g. how the columns are separated, whether there is a header with column names etc. Because it's very annoying to type out all these options every time, there is a bunch of functions that different versions of `read.table()`. All that is different between these functions is that they come with different preset options, tailored to different file types. Technically, the `read.delim()` function is the one that works for our tab-separated data, but in a lot of cases, you won't know the separator of your data before you try to read it in. For `.csv` files, I tend to start with `read.csv()` and then figure out the correct separator.

```{r}
nerd <- read.csv("exercises_solutions/data_sets/nerd_data_short.csv", sep = "\t")
```

We can see in our environment that the data now has 80 variables (columns).
And using `head()`, or data also looks more like a table:

```{r}
head(nerd)
```

The data comes with a code book, which is called `nerd_data.csv` and can be found in the `data_sets` folder.
We see that the nerd questionnaire contains 26 questions, represented in the columns `Q1` - `Q26` in the data.
There is also a personality questionnaire (`TIPI1`-`TIPI10`), a test about word definition knowledge (`VCL1` - `VCL16`) and various demographic information.
Each row represents a participant.

## Working with data frames

With the data read in correctly, we can see that we have 1,000 observations, i.e. 1,000 participants.[^shortened]

[^shortened]: The actual data set has around 15,000 participants, but for the purpose of this course, I created a smaller data set with only the first 1,000 participants.

We can see this in the environment tab, but we can also ask `R` how many rows our data frame has:

```{r}
nrow(nerd)
```

Likewise, we can ask about the number of columns:

```{r}
ncol(nerd)
```

Each column has a name, just like columns in an excel file.
We can quickly take a look at all of the column names using the function `names()`.

```{r}
names(nerd)
```

The good news is: Each column in a data frame is a vector[^2], just like the ones we've worked with before!

[^2]: Technically, a column in a data frame can also be a list, but this is rather rare and doesn't need to concern us now.

One of the most common ways[^access_columns] to access a single column in a data frame is the `$` operator.
For example, let's save the column `age` in a variable:

```{r}
age <- nerd$age
head(age, 10)
```

I am using `head()` to only show the first 10 elements in `age`, because I don't want my console (or this document) to be cluttered with thousands of numbers.
You can see that by extracting the column `age`, we ended up with a vector.

[^access_columns]: There is another way to access columns in a data frame (using `[[ ]]`), and there are scenarios where you will **need** to use it.
It is not part of the data frame chapter, but when you continue your `R` journey later, you will want to check out alternative ways to access columns in a data frame in the "optional" section at the end of this document.

You can also access any row or column in a data frame using indices, just like we did with vectors.
Only this time, we have two dimensions (rows and columns).
As with the vectors, we use square brackets (`[]`) for indices.
This time, it's two number, separated by a comma: The first number is for the rows, and the second for the columns.
Here is how we get the element in the third row and the first column in `nerd`.

```{r}
nerd[3, 5]
```

If we want to get an entire column, you just leave the rows empty:

```{r}
head(nerd[ , 3], 10)
```

(Again, I use `head()` because otherwise, thousands of numbers would be printed.)

Once again, the single column we selected has been turned into a vector.

Likewise, if we want to get a single row of our data frame, with all its columns, we just leave the columns empty.

```{r}
nerd[3, ]
```

Here, we don't receive a vector as output, but a data frame with a single row (note the column names above).

Just like with vectors, we can change values in our data frame using indexing.
Again, this will permanently overwrite our data, which we don't actually want.
This is why we'll create a copy of `nerd` with a different name that we can mess around with, while the old `nerd` data will stay untouched.
(Of course, the data file itself will still be intact, not matter what happens to the variables inside of `R`. 
So if we make a mistake, we could just load in the data again.)

```{r}
mess_around <- nerd
```

Now, let's overwrite the value in the first row and second column with 999.

```{r}
head(nerd)
```

```{r}
mess_around[1, 2] <- 999
```

```{r}
head(mess_around)
```

In principle, we can also overwrite entire rows or columns.
The only requirement is that the replacement matches the data structure.
E.g., since our column has 1,000 elements, we need to replace it with a vector that also has 1,000 elements.
Or rather, as `R` phrases it in a common error message that you receive if you get it wrong: The "number of items to replace" needs to be "a multiple of replacement length".
In other words, if we replace the 1,000 elements in our column with a vector that has a number of elements we could divide 1,000 by, the replacement vector will be repeated until it has 1,000 elements.
For example, this is what the `age` column currently looks like:

```{r}
head(mess_around$age, 10)
```

Trying to overwrite it like this won't work:

```{r error = TRUE}
mess_around$age <- c(2, 4, 6)
```

This, however, will work, because 1,000 can be divided by two.
Examine what happened to the `age` column.

```{r}
head(mess_around$age, 10)
mess_around$age <- c(2, 4)
head(mess_around$age, 10)
```
```{r}
example_vector <- c(1, 2, 4)
example_vector * c(1, 10)
```

1.1) Select the column `gender` from the `nerd` data frame and save it into a variable.

```{r}
gender <- nerd$gender
```

1.2) Select the 10th row of `nerd` and save it into a variable.

```{r}
tenth_row <- nerd[10, ]
```

1.3) Select row 10 - 13 from `nerd` and save it into a variable.

```{r}
tenth_thirteen <- nerd[10:13, ]
```

1.4) Extract the value in the 10th row and the 70th column.

```{r}
nerd[10, 70]
```

1.5) Use the "mess around data frame" to overwrite the value in the 10th row and the 70th column with a value of your choice.

```{r}
mess_around[10, 70] <- "alien science"
```

1.6) Use the "mess around data frame" to overwrite the entire 1st column with a single value of your choice.

```{r}
head(mess_around)
mess_around[ , 1] <- "something"
mess_around$Q1 <- "something"
head(mess_around)
```

### Optional

1.7) When we changed entire columns of our data frame, we took advantage of the fact that `R` is vectorised and e.g. "repeats" a single value for us so it can fill an entire column. What if we wanted to be explicit about the length of our replacement - e.g. if we want to replace the `age` column in our data frame with a vector that has 1,000 elements? Here, the `rep()` function comes in handy.

Take a look at what it does:

```{r}
rep(2, 4)
rep(c(1, 222), 3)
```
You can either specify the number of repetitions using the `each` or the `times` argument.
(If you use the second argument, without providing a name, you are using the `times` argument by default).
Observe the difference between the two:

```{r}
rep(1:3, times = 2)
rep(1:3, each = 2)
```

1.8) Can you replace the `age` column in `mess_around` with a vector where the first 500 elements are 1 and the second 500 elements are 2?

#### Create a data frame

We already know how to create vectors - here is how you create a data frame using the `data.frame()` function.

```{r}
# Note: These aren't actual recipes and I wouldn't attempt to make them
cake_recipes <- 
  data.frame(
    eggs = c(2, 4, 6, 10),
    flour = c(400, 500, 200, 100),
    butter = c(50, 150, 100, 100),
    milk = c(20, 0, 120, 50),
    baking_soda = c(rep(TRUE, 3), FALSE),
    difficulty = c("easy", "medium", "difficult", "easy")
  )

cake_recipes
```

#### Alternative ways of accessing columns

The `$` operator is probably the most common way to access a column in a data frame, because it is quick to type (due to the autocompletion feature in RStudio) and rather easy to read.
However, other approaches have advantages because they allow us to create code that is more flexible.

For example, we can use `[["column_name"]]`, i.e. the column name as a string in double square brackets.

```{r}
cake_recipes[["difficulty"]]
```

The disadvantage is that you have to type the whole column name yourself.
But the advantage is that you can use strings to access columns, and you can change these strings dynamically in your code.

```{r}
column_we_want <- "eggs"
cake_recipes[[column_we_want]]

column_we_want <- "flour"
cake_recipes[[column_we_want]]
```

The name if the column we want was stored in a variable.
When we changed the content of the variable, it gave us access to a different column.
This comes in very handy if we want to write flexible code - e.g. when we want to write a function that works on a column of which we don't know the name yet, because the user provides the column name as an argument.

Another advantage is that this approach is a little bit less "dangerous".
When using `$`, there is this feature called "partial matching".
It's a bit like autocomplete on steroids.
Even if you don't type the full column name, `R` will still identify a column that matches what you've written so far (unless there is ambiguity).
In very unfortunate cases, this can lead to silent errors (errors that don't give you an error message) when you have a typo in your column name.
I this all sounds a bit vague, here is a [tiny post with examples](https://einglasrotwein.github.io/debugging_hell/incomplete_colnames) that I've written.
This kind of problem doesn't happen when you access columns using `[[]]`, because every typo will throw an error.

### Summaries

We've used table on single vectors in the "Vectors and Functions" script.
But `table()` can also be used on several vectors at once (even more than two - try it out!).
With this code, we get an overview about how many people from which gender choose which nerdiness-category:

```{r}
table(nerd$nerdy, nerd$gender)
```

2.1) We've also used `summary()` on vectors - what happens if you use `summary()` on the whole data frame?

When we used `summary()` on the entire data frame, the output was a bit overwhelming.
The package `skimr` can help us out here, because it produces very elegant summaries.
If you want to give it a spin, you have to install it first (run the line that is currently commented out).

```{r}
# install.packages("skimr")

# Usually, I would first load the package like this: library(skimr)
# Here, I quickly use the skim() function from skimr by telling R that I'm referring
# to the skimr package using ::
# You can use any function from a package with this notation without having to
# the package first.
# skimr::skim(nerd)

# The output might be easier to read if you run it in the console.
# It works best when combined with the view function from the package "tibble".
# "tibble" is part of the tidyverse we will get to know later.
```

Because `nerd` is quite large, the summary is still a little overwhelming, but much easier to digest than the classic `summary()` output. Also note that `skim()` gives you a tiny histrogram for each column, giving you a very basic first overview about the distribution of your variables.

**THIS IS A GIT PUSH CHECKPOINT!**
