---
title: "05 The Tidyverse"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In case you still have stuff in your environment from the previous exercises, this is a reminder to clear your environment with `rm(list = ls()`.

## The tidyverse

For the next steps, I would like to introduce you to the `tidyverse`.
The `tidyverse` is an `R` package - a collection of `R` packages, in fact, that come with many handy functions for data wrangling and visualisation.
It is perfectly normal to use various `R` packages for additional functions and you will "collect" a lot of them over time.

The `tidyverse` is a little bit different, because it has its own "coding style" which varies a bit from what we've seen so far.
This has lead to a whole war between "base `R`" and "the `tidyverse`", with some hardliners on both sides.
Sometimes, the `tidyverse` is described as being especially beginner friendly and, well ... tidy.
Personally, I use both base `R` and the `tidyverse` - whatever works best for the task at hand.
They both have their pros and cons.
However, for this course, I have to make decisions about what to teach you, and especially for all things data wrangling, I will mainly show you the `tidyverse` way.

Let's install the `tidyverse` together.

Just uncomment the line below - the code is commented out so it isn't run every time you run this Rmarkdown file.

```{r eval=FALSE}
# install.packages("tidyverse")
```

We have installed our package, but we can't use its functions yet.
Packages only become "active" if you load them with the function `library()`.
(`rmarkdown` was an exception - we have been using it all this time without loading it.)
Loading a package every time you want to use it seems like a bit of effort, but actually has good reasons: Because there are so many packages, a lot of the use the same names for different functions.
One function in one package can do an entirely different thing than a function with the same name in a different package!
If you would load every package that you have installed, it would be really hard to track which functions of the same name are currently "active".
This is why you only load the packages that you need for a given script.
Once a package is loaded, it stays loaded throughout the entire `R` session.
Let's load the `tidyverse`.

```{r}
# load the tidyverse
library(tidyverse)
```

I'm a bit of a hypocrite here, because when loading the `tidyverse`, I am loading a bunch of packages, not just the ones I absolutely need for this script.
However, the `tidyverse` is intended to be used as a "package family", so it isn't so bad.
(And it's quite common not to know which specific package functions from the `tidyverse` are from.)

One of the most popular and noticeable features of the `tidyverse` is the pipe operator, which looks like this: `%>%`.[^pipe_shortcut]

[^pipe_shortcut]: If you try typing this by hand, you will lose your sanity at some point. Promise. I strongly recommend creating a shortcut. You can do this via Tools $\to$ modify keyboard shortcuts. Look for "insert pipe operator" in the search bar. I chose "alt + ." to add the pipe, simply because "alt + -" inserts the assignment operator `<-` per default. This way, I have my two most important shortcuts "next to each other".

It is used to create chains of functions, which means that you put functions after each other, linking them with the pipe operator.
In base `R`, you achieve the same thing with a more "onion-like approach".
Let me show you what I mean by this.
Let's create an example vector:

```{r}
example_vector <- c(1, 3, 33, 50, 5, 1, 2)
```

We want to take the mean of this vector (using the `mean()` function), and then round the result to two decimal places (using the function `round()`).
In base `R`, we would stack the functions into each other.[^base_pipe]
They are then executed from the inside towards the outside.

[^base_pipe]: In fact, a brand new pipe operator has just been added to base `R`, which looks like this: `|>`. It works much like the `tidyverse` pipe, but not quite. You can't exchange one for the other, at least not at the time of writing. The base `R` people is officially available since 18.05.2021 (since version 4.1.0), so it will probably face some developments during the next years.

```{r}
# base R style
round(mean(example_vector), 2)
```

In the tidyverse, we chain those functions together using the pipe operator:

```{r}
# tidyverse style
example_vector %>% mean() %>% round(2)
```

Note how we can leave out the actual data argument every time.
`mean()` doesn't use any arguments at all - the parentheses just stay empty.
That is because the pipe "delivers" `example_vector` to the `mean()` function.
That means, `mean()` **does** get an argument, only that it's not coming from within the brackets, but from the pipe to its left.
We see that `round()` still has the 2 (the number of digits we want to round to) as its second argument.
It looks like `2` is the second argument, but that's just because the first argument has been omitted.
That's again because the data - the mean of `example_vector` in this case - is coming from the left.
Let's practice chaining some functions.

Turn these layered base `R` function calls into a `tidyverse` chain.

1.1) Before you transform this one to the `tidyverse`, find out what the function `abs()` does.

```{r}
example_vector <- c(3, -2, 5, 99, -132.5)
as.character(mean(abs(example_vector)))
```

```{r}
example_vector %>% 
  abs() %>% 
  mean() %>% 
  as.character()
```

1.2) Let's make things a bit trickier with additional arguments.

```{r}
example_vector <- c(3, -2, 5, NA, 99, -132.5, NA)
as.character(mean(head(abs(example_vector), 4), na.rm = TRUE))
```

```{r}
example_vector %>% 
  abs() %>% 
  head(4) %>% 
  mean(na.rm = TRUE) %>% 
  as.character()
```

### Optional

1.3) How do we get this into a pipe? Hint: Use parentheses!

```{r}
example_vector <- c(3, -2, 5, 99, -132.5)
mean(example_vector * 2)
```

This does not work:

```{r}
example_vector * 2 %>% mean()
```

PS: Since `*` is also a function (it's just not written in the usual syntax), we can also pipe the data into `*`, with 2 as its argument. For this to work, we need to use backticks.

```{r}
example_vector %>% `*`(2) %>% mean()
```

**THIS IS A GITHUB PUSH CHECKPOINT**
