---
title: "06 Plots - Long Format"
output: 
  html_document:
    theme: dark
    highlight: zenburn
    df_print: paged
---

```{css, echo=FALSE}
.dark-output {
  background-color: #5a5a5a;
    color: white;
}

.error-output {
  background-color: #cc9393;
    color: black;
}

.message-output {
  background-color: #5a5a5a;
    color: white;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, class.output = "dark-output", class.error = "error-output",
  class.message = "message-output", fig.align = "center"
)
```

We can do tons of plots with the nerd data as it is, but in a lot of cases, our data is just not in the right format to plot it, at least when if we want to use `ggplot2`.[^we_want_ggplot]
That would e.g. be the case if we wanted do a boxplot for each of the nerdy items, `Q1` - `Q26`, next to each other.
The problem is: Each item is stored in a separate column, but we can only put **one** column on the x-axis.

[^we_want_ggplot]: And we **do** want to use `ggplot2`. Of course, you can technically achieve everything by using base `R` plotting methods (which we won't get into here), but it is usually much more tedious. `ggplot2` offers a lot of functionalities and if vanilla `ggplot2` doesn't do the job, there's loads of expansion packages that can e.g. create "publication ready plots" for you (this is e.g. how `ggpubr` is described).

Before we tackle the issue, clear your work space using `rm(list = ls())` in case you still have stuff in there from the previous exercises.
Then, load the `tidyverse` and our preprocessed nerd data:

```{r}
library(tidyverse)
source("nerd_preprocessing.R")
```

Back to our problem - we want the items 1 - 26 to be on the x-axis, which are all stored in different columns, but we can only put one column on the x-axis.
What we would need is one column, e.g. called "item", that contains the item numbers, and then another column, e.g. "score", that contains the score a participant got for the respective item.
Here's a comparison of the two formats only taking into account four participants and three items.
Don't look at the code for now, we'll get to that.
Only wrap your head around what's different in the output!

```{r}
# Wide format
nerd_red[1:4, ] %>% 
  select(id, Q1:Q3)
```

```{r}
# Long format
nerd_red[1:4, ] %>% 
  select(id, Q1:Q3) %>%
  pivot_longer(
    Q1:Q3,
    names_to = "item",
    values_to = "score"
  )
```

In the first example, each participant takes up exactly one row in the data, a very common data format.
Each item lives in a different column, which is why the data is very "wide" (if there were more items, you would need to scroll horizontally through the data).
For this reason, this format is called "wide format".

In the second example, each participant takes up several rows in the data.
Three, to be precise: One for each item.
The data aren't "wide" anymore, but instead, it got "longer" (you need to scroll horizontally), which is why this format is called the "long format".
`ggplot2` (and in fact, a lot of `tidyverse` functions) require the "long format".

You may know the transformation from wide to long format (or the other way around) other programmes like SPSS.
Within the `tidyverse`, this transformation is referred to as "pivoting" the data, which is why the function that turns "wide" data into "long" data is called `pivot_longer()`.
Let's see how it works in the example from above.
Here is the tiny portion of the nerd data (three items, four participants) we will work with:

```{r}
wide_format <- 
  nerd_red[1:4, ] %>% 
  select(id, Q1:Q3)

wide_format
```

Note how important it is that the data has an id column.
If we didn't have a participant id, it would be really hard to tell which data belongs to which participant after pivoting.

Let's put our wide data into the `pivot_longer()` function.
First, we tell `pivot_longer()` which columns should be pivoted, in our case, items 1 - 3, i.e. `Q1` - `Q3`.
You may think: Wait a minute. But the whole data frame is being transformed, aren't **all** columns part of the pivot?
In a way, that's true: The whole data frame is affected by the pivoting.
However, there are some columns that will not exist anymore after we've pivoted them (or rather, they've been transformed into a bunch of different columns).
For example, our new long data frame will not have the columns `Q1` - `Q3` anymore.
Those are the columns we specify in the `pivot_longer()` function.

The next part we need is a column where the names of our old columns go, i.e. the column that will contain the information "Q1", "Q2" and "Q3" (in our example above, we called this column "item").
We provide the name for

Next, we use the argument `names_to` to name our "naming" column.
This column will hold the old column names.
And lastly, we use the argument `values_to` to name the column that holds the values that were previously scored in the three different item columns.

`pivot_longer()` is a very powerful function that can handle much more complicated cases where multiple columns are turned into multiple naming columns, but this does not concern us for now.
However, one handy trick is to clean up the naming column a little bit:

```{r}
wide_format %>% 
  pivot_longer(
    item_1:item_3,
    names_to = "item",
    values_to = "score",
    names_prefix = "item_"
  )
```

Let's practice this a little bit.

1.2) Select the items from the Ten Item Personality Inventory from the `nerd` data (**not** `nerd_red`!). Those are named `TIPI1` - `TIPI10`. Also store the `age` column and create a column with a participant id. Save the resulting data frame into a new variable. Then, pivot the data into the long format, with a column called `TIPI_item` that stores the item name, and a column called `score` that stores the participants' response.

```{r}
# Select the age column and the items from the Ten Item Personality Inventory from the `nerd` data 
# (**not** `nerd_red`!). Those are named `TIPI1` - `TIPI10`.
# Create a column with a participant id.
# Save the resulting data frame into a new variable.
# Then, pivot the data into the long format, with a column called `TIPI_item` that stores the 
# item name, and a column called `score` that stores the participants' response.
```

1.3) Now back to our initial plan. Take `nerd_red` and transform it into the long format in such a way that you end up with a column storing the nerd item name (`Q1`- `Q26`) and another column storing the response to each item.

1.4) Create a boxplot with each nerd item on the x-axis and the score on the y-axis.

## Free solo

Now it's time to brain storm some plot ideas together.
Which plots do you want to see?
You suggest a plot, and then we will discuss together how we can build that plot.

**THIS IS A GITHUB PUSH CHECKPOINT**
