---
title: "03 Data Frames"
output: 
  html_document:
    theme: dark
    highlight: zenburn
---

```{css, echo=FALSE}
.dark-output {
  background-color: #5a5a5a;
  color: white;
}

.error-output {
  background-color: #cc9393;
  color: black;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, class.output = "dark-output", class.error = "error-output"
  )
```

Most likely, the most frequent data format you will encounter during data analysis is the data frame.
It is more or less a table of data, a bit like you know it from excel.
We can create data frames by hand, but I think it is best to get our hands on real data right from the start.
We use data from the "Nerdy Personality Attributes Scale", a fun scale that was developed to quantify what "nerdiness" is.
You can take the test online at [Open Psychometrics](https://openpsychometrics.org/tests/NPAS/).
The data for the test (and many others) is also freely available [here](https://openpsychometrics.org/_rawdata/).

The nerd data is in a .csv file - .csv and text files are really easy to read into `R`.
For excel files or SPSS data sets, we need additional packages to do the job.
The function `read.csv()` does exactly what the name suggests: It reads in a .csv file.
The first argument is the path that the file lies in.
In the folder where your script for this exercise lies, there is folder called `"data_sets"` which - you guessed it - contains the data sets for this workshop.
(It's best to keep your data in a separate folder so you won't end up in a folder where scripts, data and lord knows what else is mixed together.)
The file we want to work with is `"nerd_data.csv"`.

So, the path to our file is: `"./data_sets/nerd_data.csv"`.
This is called a **relative path** - it is the path to the file that we need to take from our current location, not the full path from the start.
The absolute path on my computer would e.g. be `"C:/Users/julit/OneDrive/Dokumente/r_stuff/r_workshop_pug/scripts_excercises/data_sets"`.
This is not only an annoyingly long path, it is also dangerous for reproducibility:
If I had put this path into the script, you could not run it, because this path does not exist on your computer.
Each of you would have to put in their own user name, the name of their workshop folder and so on.
The relative path, however, works relative to this R Markdown file.
`./"` means "here" (i.e. where the R Markdown file is), and from here, we go to the data_sets folder and then to the specific file, nerd_data.csv.
**Note that this works differently for R Markdown files versus R scripts. In a script, your location is where your current working directory is. When you're using R Projects, that's where your R Project lies.**

So, we read in our data and assign it to a variable called `nerd`.
This code runs fine, but ...

```{r}
nerd <- read.csv("./data_sets/nerd_data_short.csv")
```

... when we inspect the data using the `head()` function, it looks very weird:

```{r}
# run head() on nerd
```

If we look into our Environment tab, it says that nerd has 1,000 observations (i.e. rows) of 1 variable - i.e. one column.
However, we expected something a little bit more like a table.
This happened because `R` did not know how to separate the data.
It put all of the information it found into one single column.
We need to tell it what separates the columns.
In this data, we can see that the separator is a tab - this is indicated with `\t`.
Other common separators are `;`, `,` or `.`.
The separator argument of the `read.csv()` function is called `sep`.

```{r}
# read in nerd again, but this time with sep = "\t
```

We can see in our environment that the data now has 80 variables (columns).
And using `head()`, or data also looks more like a table:

```{r}
# run head again
```

The data comes with a code book, which is called `nerd_data.csv` and can be found in the `data_sets` folder.
We see that the nerd questionnaire contains 26 questions, represented in the columns `Q1` - `Q26` in the data.
There is also a personality questionnaire (`TIPI1`-`TIPI10`), a test about word definition knowledge (`VCL1` - `VCL16`) and various demographic information.
Each row represents a participant.

## Working with data frames

With the data read in correctly, we can see that we have 1,000 observations, i.e. 1,000 participants.[^1]

[^1]: The actual data set has around 15,000 participants, but for the purpose of this course, I created a smaller data set with only the first 1,000 participants.

We can see this in the Environment tab, but we can also ask `R` how many rows our data frame has:

```{r}
# check the number of rows of nerd
```

Likewise, we can ask about the number of columns:

```{r}
# check the number of columns of nerd
```

You might be tempted to use `length()` to check how many rows a data frame has (at least I make this mistake around once a week), but this is not what `length()` gives us for a data frame.
Do you see what `length()` does here?

```{r}
# use length on nerd
```

Before we do anything with the data, we want to familiarise ourselves with it.
`R` can help you to get an overview about your data in various ways.
Before that, however, we need to understand how data frames work.
The good news is: It's just a bunch of vectors!
Each column in a data frame is a vector[^2], just like the ones we've worked with before!

[^2]: Technically, a column in a data frame can also be a list, but this is rather rare and does not need to concern us now.

Each column has a name, just like columns in an excel file.
We can quickly take a look at all of the column names using the function `names()`.

```{r}
# check the names of nerd
```

One of the most common ways to access a single column in a data frame is the `$` operator.
Here is how it works: You type the name of the data frame, then the `$` operator, and then the name of the column you want to access.
R Studio will already suggest columns to you while you are typing.
Let's save the age in a variable:

```{r}
# save the age column into a variable
# inspect the variable using head()
```

I am using `head()` to only show the first 10 elements in `age`, because I don't want my console (or this document) to be cluttered with thousands of numbers.
Of course, you can also apply `head()` directly when using the `$` operator:

```{r}
# use head directly on the age column
```

In our console, we can see that `age` does indeed seem to be a vector.
That means, when we isolate a column of a data frame, it becomes a vector.
Remember how we assessed the type of variables in the beginning of this course?
Think about how you would confirm that `age` (or the column `nerd$age`) is a vector?

```{r}
# what is the type of the variable that we saved out age column in?
# what is the type of the age column?
# what is the type of nerd?
```

This does not really give is the answer we want to know.
But we can ask `R` directly whether something is a vector or a data frame!
The functions `is.vector()` and `is.data.frame()` do exactly what the name suggest: For a given object, they tell you whether something is a vector/data frame (`TRUE`) or not (`FALSE`).

```{r}
# check whether (the variable we saved) the age column in is a vector
# check whether nerd and the age column are a data frame
```

This general logic exists for many data types in `R`, e.g.

```{r}
# check whether something is a character
```

The `$` operator is probably the most common way to access a column in a data frame, because it is quite quick (due to the autocompletion feature) and rather easy to read.
However, other approaches have advantages when we are coding more complex things.
We will briefly look at different options, but you don't need to remember them for now.

Instead of `$`, we can use `[["column_name"]]`, i.e. the column name in double square brackets.
(Again, I use `head()` so `R` won't print the entire column, but the "raw" code would read: `nerd[["age"]]`).
The disadvantage is that you have to type the whole column name yourself, but the advantage is that it is slightly less "dangerous"[^3] and that you can exchange the column name quite flexibly with names you generated (e.g. in a function)[^4].

[^3]: Again, this doesn't concern us for now, but when using `$`, there is this feature called "partial matching". It's a bit like autocomplete on steroids. If you do not type the full column name, `R` will still identify a column that your partial name matches (that is, unless there is no ambiguity). In very unfortunate cases, this can lead to silent errors when you have a typo in your column name, i.e. errors without an error message. This does not happen when you access columns using `[[]]`, because every typo will throw an error.

[^4]: You will later see some examples how you might want to use column names that you generated somewhere else. For now, this does not concern us.

```{r}
# select the age column using [[]]
```

You can also access any row or column in a data frame using indices.
It works much like the indices that we used for single vectors, but this time, you have two dimensions (rows and columns).
As with the vectors, we use square brackets (`[]`) for indices.
This time, it's two numbers: The first for the rows and the second for the columns.
So with this code, we get the first row of `nerd`.

```{r}
# select the first row of nerd
```

And with this code, we get the third column.
Note that we leave the second position blank when we want the whole row, but that we leave the first position blank when we want the whole column.
(Again, I use `head()` because otherwise, thousands of numbers would be printed.)

```{r}
# select the third column of nerd
```

Once again, selecting a single column will give us a vector.
But as you saw above, selecting a single row will give us a data frame with column names.

We can also extract single values in our data frame by providing an index for both row **and** column.
For example, we can select the number at the third row and the first column like this:

```{r}
# select the value in the third row and the first column
```

We can also change values in our data frame in the same manner by assigning new values to them.
**Careful, this will overwrite the values in our data frame!**
So we want to save our `nerd` data into a new variable, so we can mess around with that one instead of destroying the data in our old variable.
(Of course, the data file itself would still be intact, so we could also just load in the data again if we make a mistake.)

```{r}
# Save the nerd data into a new variable that you can mess around with
```

Now, let's overwrite the value in the first row and second column with 999.

```{r}
# Overwrite the value in the first row and second column with 999.
```

1.1) Select the column `gender` from the `nerd` data frame and save it into a variable. Use three different approaches.

1.2) Select the 10th row of `nerd` and save it into a variable.

1.3) What would you do if you wanted to select row 10, 11 and 12 from `nerd`? Think back to how we used indices with vectors.

1.4) Extract the value in the 10th row and the 70th column.

1.5) Use the "mess around data frame" to overwrite the value in the 10th row and the 70th column with a value of your choice.

1.6) Use the "mess around data frame" to overwrite the entire 70th column. We will solve this task together with the course, but first think about the following questions: How would you, in principle, overwrite the entire 70th column? What would you need to replace it with?

## Getting an overview

When we worked with vectors (the hot dog data), we already used some functions that can help us to get an overview over the data.
One of them was `summary()`.

2.1) Use `summary()` on the `age` column. What do you notice?

2.2) Use `summary()` on the `country` column. What is different here? Why? (Tip: Inspect `nerd$country`, maybe using `head()`).

2.3) Try `table()` for the `country` column. What is the difference?

2.4) What happens if you use `summary()` on the whole data frame?

2.5) Try `table()` on the gender column. Does this help? Can you find more information in the codebook?

**THIS IS A GIT PUSH CHECKPOINT!**
