---
title: "05 Plots"
output: 
  html_document:
    theme: dark
    highlight: zenburn
    df_print: paged
---

```{css, echo=FALSE}
.dark-output {
  background-color: #5a5a5a;
    color: white;
}

.error-output {
  background-color: #cc9393;
    color: black;
}

.message-output {
  background-color: #5a5a5a;
    color: white;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, class.output = "dark-output", class.error = "error-output",
  class.message = "message-output", fig.align = "center"
)
```

## Now for some fun

There is more to be said about data wrangling, and we will get back to it, but I want to create some plots first.
Plots are fun because they give you nice and colourful results, and they are a very important tool for inspecting your data and getting an intuition for it.
While base `R` has a plotting function, the `tidyverse`'s `ggplot2` is what is partly responsible fpr `R`'s fame.

`ggplot2` is based on the ["Grammar of Graphics" by Leland Wilkinson](https://www.springer.com/de/book/9780387245447), a framework for good data visualisation in data science.
In a way, `ggplot2` has a grammar of its own, which is something new for us.
But don't worry, it all makes sense.

First, we need some data.
I want to use the opportunity to show you how you would preprocess your data in a separate file, and then use it in you R Markdown file.
We want to set up a little pre-processing file for our nerd data.
To achieve this, we will go though the file `04_data_wrangling.Rmd` and copy-paste the bits of code that we need in order to get the raw data ready to be plotted.

```{r}
library(tidyverse)
# TO DO: create nerd preprocessing file!
source("nerd_preprocessing.R")
```

One basic plot that you will frequently need is a histrogram.
It's a good way to quickly inspect the distribution of your data.
Below, you see how a simple histogram for the participant age works.
The concept is pretty simple: We pipe our data, `nerd_red` into the ggplot function.
Within the function, we use the `aes()` function, which stands for "aesthetics".
The aesthetic function is used to tell `ggplot` all the information that is related to the data (as opposed to things that have nothing to do with the data, e.g. overall plot colour).
It will make more sense once we see which things belong into `aes()` and which things belong outside of it.
Unsurprisingly, the argument `x` is what we want to put on the x-axis, which is the age of our participants in this case.
Lastly, we tell ggplot to make a histogram.
Ggplot uses different "geoms", which you can imagine as different "geometric" plot shapes.
Other geoms are for example `geom_point`, `geom_bar` or `geom_boxplot`.
You can probably guess what they do.
The special thing is that we don't use a pipe to add the histogram, but a plus sign.
That is because ggplots work in "layers".
You might have used an image maniplation program (or something like PowerPoint) that followed the same logic: When you created an image, some elements were behind/in front of each other.
In ggplot, that's the same.
We have the plot background as first layer, and then there's our histrogram on top of that.

```{r}
# plot a histogram with age on the x axis
```

We can observe the layered structure by leaving out the histogram layer.
Here, we can see that ggplot still drew the plot background for us.
It also labelled the x-axis with "age", because it knows that our age data will go there.
Furthermore, it picked a sensible scale for the axis, because it knows the minimum/maximum of our data and can figure out the range it will need to plot.
However, no data is plotted, and there is no y-axis.
We did not provide any information for the y-axis.
In our previous plot, `geom_histogram` automatically counted how many participants were in each age bin, and based the y-axis on that.

```{r}
# just use ggplot() with age on the x axis, without a geom
```

Let's check out a different geom.
Try `geom_boxplot()`.
It looks a bit weird - what might be the problem here?

```{r}
# Try geom_boxplot instead of geom_histogram
```

The plot will look much better with a different orientation, i.e. the data on the y-axis.

```{r}
# Fix the problem from above
```

But still, the x-axis looks weird.
Here's how we can fix that.
There's still that nasty little x as axis label - we'll talk about how to get rid of that later.

```{r}
# use a dummy ("") for the x-axis
```

For now, let's try another geom.
How about a good old scatter plot?
However, this requires information about both x and y, which makes sense.
What do we want to plot on next to age?
Maybe the nerd score?

```{r}
# scatter plot with age in the x axis and nerd score on the y axis
```

How about adding a line to the data, indicating how age and nerd score are associated with each other?
The corresponding geom is called `geom_smooth`.

```{r}
# add a geom_smooth
```

The default method for `geom_smooth`, as `R` informs us, is "loess".
We can also have a plain linear regression line, which we achieve like this:

```{r}
# use a linear model instead
```

We can also split our plot by another variable, e.g. gender.
Gender is a property of our data, so it also goes into the `aes()` function.

```{r}
# same plot as above, but colour by gender
```

## Customise things

This is all fun and games, but the plot is not as nice as it could be.
We want to customise it a little bit.
For example, the text on the y axis says "nerd_score", because ggplot just put in the name of the column.
We can change that in the following way:

```{r}
# customise the y-axis label
```

Next, let's move the legend to the top of the plot.
Using `theme()`, we can control **a lot** of things in our plot, like font sizes, whether there should be grid lines or not and much, much more.

```{r}
# put the legend on top
```

This grey plot grid is something ggplot is very famous for.
However, it might not be the most beautiful and modern plotting option.
You can either set up a completely custom theme using the `theme()` function, or you can use one of the many default themes, for example:

```{r}
# pick a different theme (theme_bw)
```

Much better. How about more fine-grained steps on the x and y axis?
One quick way to set up the breaks on a plot axis is the function `seq()`.
It generates a sequence of numbers.
You give it the start point, the end point, and the stepsize:

```{r}
# create a sequence from 1 - 10 in steps of 2
```

It is also possible not to give a step size, but the **number** of steps instead:

```{r}
# create steps from 1 - 10 with length.out = 3
```

So, we can use `seq()` to adjust our axes like this:

```{r}
# adjust the steps of the x and y axis
```

How about some thicker lines for the regression lines?
And maybe we should make the dots and the confidence interval a bit more opaque.
Note that all of these things don't go inside the `aes()` function, because things like transparency or line size don't depend on the data.

```{r}
# use alpha and size outside of the aes function
```

To illustrate this, let's make our first histogram a little bit more beautiful.
Here is what it looked like before:

```{r}
# remember what our first histogram looked like
```

Let's change the theme and also add some colour.
Note that some for some geoms or shapes, there is a distinction between colour and fill.
Fill does exactly what the name suggests, and colour refers to the colour of the border.
The circles we used before did not have a border, so we could colour the complete circle using colour.
We can refer to the colour that we want by name, but can also use hex codes.

```{r}
# change the fill to"#dca3a3" and the colour to black
```

How about categorical data?
Let's have a boxplot for age that is grouped by gender.
Here, I use gender on both the x-axis and for the colour.
That's a bit redundant, but we can hide the x-axis if we want, so only the legend carries the information.
I have also added a title.

```{r}
# create a box plot with y = age and x = gender.
# fill by gender
# set a title
# use a nice theme and move the legend to the top
```

How about adding another factor?

1.1) Can you do a boxplot that has the nerd score on the y-axis, gender on the x-axis and is filled with whether the person voted or not?

## Detour: Long format

What if we wanted do a boxplot for each of the nerdy items, `Q1` - `Q2` next to each other?
We run into a problem: Each item is stored in a separate column, but we can only put **one** column on the x-axis.
That is because `ggplot2` (and in fact, a lot of `tidyverse` functions) require the "long format".
The long format means that we deviate from the classical data format where every participant has a single row.
In the long format, one participant can take up several rows.
Consider this simplified example of a data frame with two participants and three items:

```{r}
wide_format <- 
  data.frame(
    participant = 1:2,
    item_1 = c(1, 4),
    item_2 = c(5, 10),
    item_3 = c(11, 2)
  )

wide_format
```

We want to "pivot" the data, so we end up with a situation like this (only observe the output for now, not the code):

```{r}
wide_format %>% 
  pivot_longer(
    item_1:item_3,
    names_to = "item",
    values_to = "score"
  )
```

As you can see, each participant now takes up three rows (one for each item), and the corresponding score is stored in a single, separate column.
This would now e.g. allow us to build a plot where we put the item on the x-axis and the score on the y-axis.

You may know this transformation from other programmes like SPSS.
In the `tidyverse`, we achieve this by using the `pivot_longer()` function.
Here is how it works: First, we tell it which columns should be part of the pivot, in our case, `item1` - `item3`.
You may think: Wait a minute. But the whole data frame is being transformed, aren't all columns part of the pivot?
Think of it like this: We provide the columns that should be gone in the resulting "long" data frame, and replaced by other columns.
In our case, we want to get rid of the individual item columns, but turn them into a column coding which item we have and a corresponding score column.
Next, we use the argument `names_to` to name our "naming" column.
This column will hold the old column names.
And lastly, we use the argument `values_to` to name the column that holds the values that were previously scored in the three different item columns.

`pivot_longer()` is a very powerful function that can handle much more complicated cases where multiple columns are turned into multiple naming columns, but this does not concern us for now.
However, one handy trick is to clean up the naming column a little bit:

```{r}
wide_format %>% 
  pivot_longer(
    item_1:item_3,
    names_to = "item",
    values_to = "score",
    names_prefix = "item_"
  )
```

Let's practice this a little bit.

1.2) Select the items from the Ten Item Personality Inventory from the `nerd` data (**not** `nerd_red`!). Those are named `TIPI1` - `TIPI10`. Also store the `age` column and create a column with a participant id. Save the resulting data frame into a new variable. Then, pivot the data into the long format, with a column called `TIPI_item` that stores the item name, and a column called `score` that stores the participants' response.

```{r}
# Select the age column and the items from the Ten Item Personality Inventory from the `nerd` data 
# (**not** `nerd_red`!). Those are named `TIPI1` - `TIPI10`.
# Create a column with a participant id.
# Save the resulting data frame into a new variable.
# Then, pivot the data into the long format, with a column called `TIPI_item` that stores the 
# item name, and a column called `score` that stores the participants' response.
```

1.3) Now back to our initial plan. Take `nerd_red` and transform it into the long format in such a way that you end up with a column storing the nerd item name (`Q1`- `Q26`) and another column storing the response to each item.

1.4) Create a boxplot with each nerd item on the x-axis and the score on the y-axis.

## Free solo

Now it's time to brain storm some plot ideas together.
Which plots do you want to see?
You suggest a plot, and then we will discuss together how we can build that plot.

**THIS IS A GITHUB PUSH CHECKPOINT**
